Fast and Scalable Mixed Precision Euclidean Distance Calculations Using GPU Tensor Cores

### 2025-07-28 Paper/main.tex
Removed tabularx package.
Deleted column definitions, and had to manually call out some column
widths using p{size}. Had to add ragged formatting some places as well
to fix the formatting.
Removed packages already in acmart package.
Shrunk some tables and graphics to eliminate warnings.
Was getting some overfull \hbox warnings and was able to easily adjust
these to fix them.
Reviewed paper and made small fixes.

%% End of template, start of our custom Preamble here:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%custom commands
\usepackage[utf8]{inputenc}
\usepackage{colortbl}
\usepackage{amsmath,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[dvipsnames]{xcolor}
\usepackage{enumitem}
\usepackage{subfigure}
\usepackage{subcaption}

\usepackage{listings}
\lstset{
language=C++,
basicstyle=\small,
showstringspaces=false,
}
\usepackage{bm}

\usepackage{enumitem}
\setlist{topsep=0pt, leftmargin=*}

\usepackage{xcolor}
\usepackage{xparse}

\usepackage{xparse}

\usepackage{xspace}    
\usepackage{tabularx}

%centering tabularx
\newcolumntype{Y}{>{\centering\arraybackslash}X}
%right align in tabularx
\newcolumntype{R}{>{\raggedleft\arraybackslash}X}

%right align in tabularx
\newcolumntype{L}{>{\raggedright\arraybackslash}X}

@@

\definecolor{OliveGreen}{cmyk}{0.64,0,0.95,0.40}
\definecolor{CadetBlue}{cmyk}{0.62,0.57,0.23,0}
\definecolor{lightlightgray}{gray}{0.93}
\lstset{
language=C++,                          % Code langugage
basicstyle=\ttfamily,                   % Code font, Examples: \footnotesize, \ttfamily
keywordstyle=\color{OliveGreen},        % Keywords font ('*' = uppercase)
commentstyle=\color{gray},              % Comments font
numbers=left,                           % Line nums position
numberstyle=\tiny,                      % Line-numbers fonts
stepnumber=1,                           % Step between two line-numbers
numbersep=5pt,                          % How far are line-numbers from code
backgroundcolor=\color{lightlightgray}, % Choose background color
frame=none,                             % A frame around the code
tabsize=2,                              % Default tab size
captionpos=t,                           % Caption-position = bottom
breaklines=true,                        % Automatic line breaking?
breakatwhitespace=false,                % Automatic breaks only at whitespace?
showspaces=false,                       % Dont make spaces visible
showtabs=false,                         % Dont make tabls visible
columns=flexible,                       % Column format
morekeywords={__global__, __device__},  % CUDA specific keywords
}

\usepackage{framed}

\definecolor{TFFrameColor}{rgb}{0.0,0.0,0.5} % darkish blue frame
\definecolor{TFBackgroundColor}{rgb}{0.9,0.95,1} % light blue background
\definecolor{TFTitleColor}{rgb}{1,1,1} % white text

@@

\begin{abstract}
Modern GPUs are equipped with tensor cores (TCs) that are commonly used for matrix multiplication in artificial intelligence workloads. However, because they have high computational throughput, they can lead to significant performance gains in other algorithms if they can be successfully exploited. We examine using TCs to compute Euclidean distance calculations, which are used in many data analytics applications. Prior work has only investigated using 64~bit floating point (FP64) data for computation; however, TCs can operate on lower precision floating point data (i.e., 16 bit matrix multiplication and 32 bit accumulation), which we refer to as FP16-32. FP16-32 TC peak throughput is so high that TCs are easily starved of data.  We propose a Fast and Scalable Tensor core Euclidean Distance (FaSTED) algorithm. To achieve high computational throughput, we design  \ouralg for significant hierarchical reuse of data and maximize memory utilization at every level (global memory, shared memory, and registers).  We apply \ouralg to the application of similarity searches, which typically employ an indexing data structure to eliminate superfluous Euclidean distance calculations. We compare to the state-of-the-art (SOTA) TC Euclidean distance algorithm in the literature that employs FP64, as well as to two single precision (FP32) CUDA core algorithms that both employ an index. We find that across four real-world high-dimensional datasets spanning 128-960 dimensions, the mixed-precision brute force approach achieves a speedup over the SOTA algorithms of 2.5--51$\times$. We also quantify the accuracy loss of our mixed precision algorithm to be $<$0.06\% when compared to the FP64 baseline.

@@

In this paper, we present a Fast and Scalable Tensor core Euclidean Distance (\ouralg) algorithm, that is designed for moderate to high-dimensional datasets.\footnote{The source code is publicly available: \url{https://github.com/bwcurless/FaSTED}.} We consider comparing all points in an input dataset to each other, which has a quadratic time complexity, and only return those points that are within a search distance, $\epsilon$, of each other. This is a common application scenario, as it is assumed in many of the data analytics applications described above. \ouralg can be employed as a subroutine in algorithms and workloads requiring the large scale computation of Euclidean distances.

@@

{
\addtolength{\belowdisplayskip}{-1.0ex}
\addtolength{\belowdisplayshortskip}{-0.5ex}
\addtolength{\abovedisplayskip}{-2.0ex}
\addtolength{\abovedisplayshortskip}{-2.0ex}
% \begin{equation}
% dist(p_i, p_j) = \sqrt{\sum^{d}_{k=1}(p_{i,k} - p_{j,k})^2}= \sqrt{\sum^{d}_{k=1}p_{i,k}^2 - 2p_ip_j + p_{j,k}^2}.
% \label{eqn:euclidean_distance}
% \end{equation}
\begin{equation}
dist(p_i, p_j) = \sqrt{\sum^{d}_{k=1}(p_{i,k} - p_{j,k})^2}= \sqrt{\sum^{d}_{k=1}p_{i,k}^2 - 2p_{i,k}p_{j,k} + p_{j,k}^2}.
\label{eqn:euclidean_distance}
\end{equation}
}

@@

\begin{table}[]

\footnotesize
    \centering
    \caption{Comparison of matrix sizes for FP16-32.}
        \begin{tabularx}{\columnwidth}{l|L|L} 
        \hline
        \rowcolor{blue!40!white}Size ($m$-$n$-$k$)&WMMA API&\ptxmma\\\hline
        16$\times$16$\times$16&\checkmark&\\
        32$\times$8$\times$16&\checkmark&\\
        8$\times$32$\times$16&\checkmark&\\
        8$\times$8$\times$4&&\checkmark\\
        16$\times$8$\times$8&&\checkmark\\
        \textbf{16$\times$8$\times$16} (Used by \ouralg)  &&\checkmark\\
        \hline
        \end{tabularx}
    \label{tab:WMMA_vs_PTX}
\end{table}

@@

There are two main PTX instructions for using TCs. The first is \texttt{ldmatrix}, shown in Listing~\ref{lst:ldmatrix}, which creates a fragment of a matrix that is stored in registers shared across a warp. The second is \texttt{mma.sync}, shown in Listing~\ref{lst:mma}, which computes $\bm{D}=\bm{A}\times \bm{B}+\bm{C}$, where $\bm{A}$, $\bm{B}$, $\bm{C}$, and $\bm{D}$ are each fragments that were created by prior \texttt{ldmatrix} instructions. %The fragment $\bm{D}$ can be substituted for $\bm{C}$ to accumulate the result of the multiplication in place. 
When using FP16-32 instructions, the data stored in matrices $\bm{A}$ and $\bm{B}$ are represented in FP16 as 16$\times$16 and 16$\times$8 fragments while the accumulator matrices, $\bm{C}$ and $\bm{D}$, are represented in FP32 as 16$\times$8 fragments.

@@

\noindent
{\bf Step~2:} Tensor cores operate on small matrices that are stored in registers, we refer to these matrices as ``register fragments,'' or ``rf''. There are three register fragments required to compute distances:
\begin{enumerate}[wide=0pt]
\item\prf: 16 dimensions from 16 points in dataset $D$. We refer to these simply as the ``points''. 
\item\qrf: 16 dimensions from 8 points in dataset $D$ in a transposed layout. We refer to these as the ``query points''. 
\item\arf: A $16\times 8$ fragment initialized with zeros.
 \end{enumerate}
 We use TCs to multiply and accumulate these three fragments to obtain: $\arf = \prf \times \qrf + \arf$. As TCs can only compute inner products of $16$ dimensions per operation (see Table~\ref{tab:WMMA_vs_PTX}), we must iteratively load new point data into \prf\ and \qrf\ and accumulate into \arf\ $\frac{d}{16}$ times.

@@

\begin{figure}
    \centering
    \includegraphics{tikz_build/Block_Tile}
    \caption{Two iterations of a $64{\times }64{\times }64$ block tile are shown. Two \sixfourd block fragments, \pbf \ and \qbf, are paged from global memory to shared memory (SMEM). Four warp tiles (Figure~\ref{fig:warp_mma}) each page a portion of the block fragments into their own register fragments, and accumulate the distances. As $d{=}128$, this sequence is repeated for two iterations.}

@@

\subsubsection{L1 Cache elimination}\label{sec:L1_elimination}
Each SM has 192~KB of memory that can be used as an L1 cache, or directly manipulated as shared memory. In order to maximize throughput, most of the L1 cache has been configured as shared memory to allow us to use a pipelined execution that hides memory access latency, which will be described in the following sections.

@@

Figure~\ref{fig:shared_memory} shows how each \eightd slice in a group of 8 threads will be stored in a unique bank, and is thus free of bank conflicts.

While swizzling the data layout is not required to have coalesced and bank conflict-free shared memory stores, it is imperative for loading the point data from shared memory and storing it into registers. To load a single fragment \prf, there are four phases, or memory transactions, in a \texttt{ldmatrix} instruction (see Listing~\ref{lst:ldmatrix} and Figure~\ref{fig:ldmatrix}). Each phase generates a single 128~B memory transaction that must be bank conflict-free to achieve maximum throughput.

To see that banks are accessed in a conflict-free manner, observe that each \eightd slice of data being accessed in a single phase in Figure~\ref{fig:ldmatrix} is located in a unique bank in Figure~\ref{fig:shared_memory}. In contrast, a simple row-major copy from global memory (Figure~\ref{fig:global_memory} into shared memory would create 8-way bank conflicts during each phase.

@@

We examine two main metrics when measuring performance as follows. \emph{Response time:} We measure the response time of the algorithms. \emph{Derived TFLOPS:} We report the TFLOPS yielded by the TCs by taking the total number of operations and dividing it by the measured response time. We refer to this when we discuss TFLOPS.

@@

To make a fair comparison between all implementations, we include all overheads, such as allocating memory, transferring data to and from the GPU and storing the results in main memory. Regarding methods that employ indexing, we include all index construction overheads.

@@

\noindent\textbf{\ouralg --} We configure \ouralg using the parameters outlined in Table~\ref{tab:optimized_parameters}, which were motivated in Section~\ref{sec:our_algorithm}.

\begin{table}[!t]
\footnotesize
\caption{Summary of optimized parameters. }
        \begin{tabularx}{\columnwidth}{L|r} 
        \hline
        \rowcolor{blue!40!white}Parameter & Optimized Value \\ \hline
        Block tile dispatch shape & $8{\times }8$ blocks \\ \hline
        Block tile iteration size & $128{\times }128{\times }64$ \\ \hline
        Number of blocks in grid& $2{\times}$ \# of SMs (216 total) \\ \hline
        Warp tile iteration size & $64{\times }64{\times }16$ \\ \hline
        Warps per block & 4 \\ \hline
        Pipeline depth & 2 \\ \hline
        \end{tabularx}
    \label{tab:optimized_parameters}
\end{table}

\noindent\textbf{\tedjoin --} As described in Section~\ref{sec:background_tedjoin}, \tedjoin~\cite{gallet2022leveraging} is the only other algorithm in the literature that computes Euclidean distances on TCs, and it uses double precision floating point (FP64) values.

@@

\noindent\textbf{\gdsjoin --} As described in Section~\ref{sec:related_work_gpu_similarity_searches} this algorithm performs index-supported range queries using CUDA cores. We use  the same optimizations and configuration as described in the \gdsjoin papers~\cite{gowanlock2019gpu,gowanlock2023optimization}, so we only outline the differences here. We configure \gdsjoin to use FP32 data, which provides a baseline of comparison to \ouralg that does not use FP64 like \tedjoin. Lastly, we use a batch size of $b_s=2\times10^9$ for batching the result set from the GPU to the host. We also employ \gdsjoin as executed using FP64 to compute the accuracy of \ouralg's mixed-precision results. The code is publicly available.\footnote{\url{https://github.com/mgowanlock/gpu_self_join}}

\noindent\textbf{\mistic --} This algorithm is similar to \gdsjoin above as it uses CUDA cores (Section~\ref{sec:related_work_gpu_similarity_searches}). We use the same configurations and optimizations as outlined in the \mistic paper~\cite{donnelly2024multi} and use a block size of 256 with 1024 blocks per kernel invocation (multiple invocations are used to batch the result set from device to host). The algorithm uses 6 levels and selects between 38 candidate layers at each level when incrementally constructing the index. \mistic is configured to use FP32 instead of FP64 (the latter was used in the evaluation in the \mistic paper~\cite{donnelly2024multi}). The code is publicly available.\footnote{\url{https://github.com/bwd29/self-join-MiSTIC}}

\noindent\textbf{Short circuiting of distance calculations -- }The CUDA core algorithms (\gdsjoin and \mistic) abort computing the distance between each pair of points when the running total distance exceeds $\epsilon$. When points in a dataset are very spread out this saves significant computation time. The TC algorithms (\tedjoin and \ouralg) are less flexible because they compute the distances between all permutations of two sets of points simultaneously. \ouralg compares an entire 128$\times$128 block of points at a time. As \ouralg is computing 16,384 distances at a time, and every single point would need to short circuit for the optimization to be viable, we forego this optimization. \tedjoin compares an 8$\times$8 matrix of points at a time in FP64 mode which is small enough that it does employ short circuiting. \tedjoinbrute does not abort early.

\begin{table}[!t]
\caption{Comparison of implementation properties.}
\footnotesize
    \centering
    {
        % Get Ted-Join-Brute to fit on one line
        % \setlength{\tabcolsep}{4pt} % default is 6pt
        \begin{tabularx}{\columnwidth}{l|l|l|L|L} 
        \hline
        \rowcolor{blue!40!white}Implementation&GPU Core&Precision& Scenario~1: Brute Force  (Sec.~\ref{sec:scenario1_brute_with_epsilon})&Scenario~2: Index-Supported (Sec.~\ref{sec:scenario2_index})\\ \hline
        \ouralg&Tensor&FP16-32&\checkmark&\\
        \tedjoinbrute&Tensor&FP64&\checkmark&\\
        \tedjoinindex&Tensor&FP64&&\checkmark\\
        \gdsjoin&CUDA&FP32&&\checkmark\\
        \mistic & CUDA & FP32 & &\checkmark\\
        \hline
        \end{tabularx}
    }
    \label{tab:implementations}
\end{table}

@@

\noindent\textbf{Selectivity of the Range Queries --} The search radius, $\epsilon$, used in a similarity search will impact the degree of pruning afforded by an indexing data structure. A large search radius that returns a large number of neighbors per point in a dataset will reduce the degree of index pruning, and a small value of $\epsilon$ will allow many points to be ignored when performing distance calculations. To standardize experiments across datasets, we employ the selectivity, which refers to the mean number of points found by each point searched in the dataset. The selectivity is defined as $S{=}(|R|{-}|D|)/|D|$, where $|R|$ is the total result set size (the number of pairs of points found within $\epsilon$ of each other). Selecting common selectivity values to use across datasets ensures that searches have a proportional amount of work to each other; otherwise, comparing performance between datasets is less meaningful. In our evaluation, we select values of $\epsilon$ for each dataset that yield three levels of selectivity (small, medium, and large), and these are defined as $S_s{=}64$, $S_m{=}128$, and $S_l{=}256$. The $\epsilon$ values corresponding to these selectivity values are given for the real-world datasets in Table~\ref{tab:datasets}.

\begin{table}[!t]
\caption{\emph{Top:} Real-world datasets that we use in our evaluation. The dataset size, $|D|$, and data dimensionality, $d$, are shown, in addition to the $\epsilon$ values that correspond to three target selectivity values that we employ ($S_{s}$,  $S_{m}$, and $S_{l}$). \emph{Bottom:} Synthetic datasets that we use in our evaluation.}
\footnotesize
    \centering
        \begin{tabularx}{\columnwidth}{L|r|r|r|r|r} 
        \hline

            \rowcolor{blue!40!white} \multicolumn{6}{c}{Real-World Datasets}\\\hline
           \rowcolor{blue!40!white}  Dataset& $|D|$ & $d$  & $\epsilon (S_{s})$  &$\epsilon (S_{m})$  &$\epsilon (S_{l})$ \\\hline
            \sift&  10,000,000 &  128 & 122.5 & 136.5 & 152.5 \\ \hline
             \tinyds&  5,000,000 &  384 & 0.1831 & 0.2045 & 0.2275 \\ \hline
             \cifar&  60,000 &  512 & 0.6289 & 0.6591 & 0.6914 \\ \hline
             \gist&  1,000,000 &  960 & 0.4736 & 0.5292 & 0.5937 \\ \hline
           \rowcolor{blue!40!white}  \multicolumn{6}{c}{Synthetic Datasets}\\\hline
          \rowcolor{blue!40!white}   Dataset& \multicolumn{2}{c|}{ $|D|$ }& \multicolumn{3}{c}{$d$}\\\hline
             \synth&\multicolumn{2}{c|}{$10^{3 + n/3} \text{ where } n\in [0, 1, \ldots, 9]$}&\multicolumn{3}{c}{$d\in 2^n \text{ where } n\in[6, 7, \ldots, 12]$}\\\hline
        \end{tabularx}
    \label{tab:datasets}
\end{table}

@@

Recall that because \ouralg is brute force, its performance is not impacted by the data distribution, so these results apply to any real-world dataset of the same size and dimensionality. To estimate \ouralg's performance on a dataset with a dimensionality that is not a multiple of $64$, one must use the throughput measurements from the next multiple of $64$. In other words, for low dimensionality like $d{=}65$, $63$ dimensions of zeros are added, and the throughput is almost halved. This effect on performance is decreases as dimensionality increases.

@@

Table~\ref{tab:accuracy} presents the Overlap Between Result Sets metric across all real-world datasets, where we find that there is a negligible accuracy loss. The maximum accuracy loss occurs on \cifar ($S_l=256$) which achieves a very high accuracy of 99.946\%.

>>>>

%% End of template, start of our custom Preamble here:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%custom commands
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{bm}
\usepackage{colortbl}
\usepackage{enumitem}
\setlist{topsep=0pt, leftmargin=*}
\usepackage{framed}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{multirow}
\usepackage{textcomp}
\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{xspace}

@@

\definecolor{OliveGreen}{cmyk}{0.64,0,0.95,0.40}
\definecolor{CadetBlue}{cmyk}{0.62,0.57,0.23,0}
\definecolor{lightlightgray}{gray}{0.93}
\lstset{
language=C++,                          % Code langugage
basicstyle=\ttfamily,                   % Code font, Examples: \footnotesize, \ttfamily
keywordstyle=\color{OliveGreen},        % Keywords font ('*' = uppercase)
commentstyle=\color{gray},              % Comments font
numbers=left,                           % Line nums position
numberstyle=\tiny,                      % Line-numbers fonts
stepnumber=1,                           % Step between two line-numbers
numbersep=5pt,                          % How far are line-numbers from code
backgroundcolor=\color{lightlightgray}, % Choose background color
frame=none,                             % A frame around the code
tabsize=2,                              % Default tab size
captionpos=t,                           % Caption-position = bottom
breaklines=true,                        % Automatic line breaking?
breakatwhitespace=false,                % Automatic breaks only at whitespace?
showspaces=false,                       % Dont make spaces visible
showstringspaces=false,                 % Dont make spaces visible inside of strings
showtabs=false,                         % Dont make tabls visible
columns=flexible,                       % Column format
morekeywords={__global__, __device__},  % CUDA specific keywords
}

% Colorbox colors for titled-frame
\definecolor{TFFrameColor}{rgb}{0.0,0.0,0.5} % darkish blue frame
\definecolor{TFBackgroundColor}{rgb}{0.9,0.95,1} % light blue background
\definecolor{TFTitleColor}{rgb}{1,1,1} % white text

@@

\begin{abstract}
Modern GPUs are equipped with tensor cores (TCs) that are commonly used for matrix multiplication in artificial intelligence workloads. However, because they have high computational throughput, they can lead to significant performance gains in other algorithms if they can be successfully exploited. We examine using TCs to compute Euclidean distance calculations, which are used in many data analytics applications. Prior work has only investigated using 64~bit floating point (FP64) data for computation; however, TCs can operate on lower precision floating point data (i.e., 16 bit matrix multiplication and 32 bit accumulation), which we refer to as FP16-32. FP16-32 TC peak throughput is so high that TCs are easily starved of data.  We propose a Fast and Scalable Tensor core Euclidean Distance (\ouralg) algorithm. To achieve high computational throughput, we design  \ouralg for significant hierarchical reuse of data and maximize memory utilization at every level (global memory, shared memory, and registers).  We apply \ouralg to the application of similarity searches, which typically employ an indexing data structure to eliminate superfluous Euclidean distance calculations. We compare to the state-of-the-art (SOTA) TC Euclidean distance algorithm in the literature that employs FP64, as well as to two single precision (FP32) CUDA core algorithms that both employ an index. We find that across four real-world high-dimensional datasets spanning 128-960 dimensions, the mixed-precision brute force approach achieves a speedup over the SOTA algorithms of 2.5--51$\times$. We also quantify the accuracy loss of our mixed precision algorithm to be $<$0.06\% when compared to the FP64 baseline.

@@

In this paper, we present a Fast and Scalable Tensor core Euclidean Distance (\ouralg) algorithm, that is designed for moderate to high-dimensional datasets.\footnote{The source code is publicly available: \url{https://github.com/bwcurless/fasted}.} We consider comparing all points in an input dataset to each other, which has a quadratic time complexity, and only returning those points that are within a search distance, $\epsilon$, of each other. This is a common application scenario, as it is assumed in many of the data analytics applications described above. \ouralg can be employed as a subroutine in algorithms and workloads requiring the large scale computation of Euclidean distances.

@@

{
\addtolength{\belowdisplayskip}{-1.0ex}
\addtolength{\belowdisplayshortskip}{-0.5ex}
\addtolength{\abovedisplayskip}{-2.0ex}
\addtolength{\abovedisplayshortskip}{-2.0ex}
% \begin{equation}
% dist(p_i, p_j) = \sqrt{\sum^{d}_{k=1}(p_{i,k} - p_{j,k})^2}= \sqrt{\sum^{d}_{k=1}p_{i,k}^2 - 2p_ip_j + p_{j,k}^2}.
% \label{eqn:euclidean_distance}
% \end{equation}
\begin{equation}
	\begin{aligned}
		dist(p_i, p_j) &= \sqrt{\sum^{d}_{k=1}(p_{i,k} - p_{j,k})^2} \\
									 &= \sqrt{\sum^{d}_{k=1}p_{i,k}^2 - 2p_{i,k}p_{j,k} + p_{j,k}^2}.
	\end{aligned}
\label{eqn:euclidean_distance}
\end{equation}
}

@@

\begin{table}[!t]

\footnotesize
    \centering
		\caption{Comparison of matrix sizes for FP16-32.}
		\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}|p{2.2cm}|p{2.2cm}} 
        \hline
        \rowcolor{blue!40!white}Size ($m$-$n$-$k$)&WMMA API&\ptxmma\\\hline
        16$\times$16$\times$16&\checkmark&\\
        32$\times$8$\times$16&\checkmark&\\
        8$\times$32$\times$16&\checkmark&\\
        8$\times$8$\times$4&&\checkmark\\
        16$\times$8$\times$8&&\checkmark\\
        \textbf{16$\times$8$\times$16} (Used by \ouralg)  &&\checkmark\\
        \hline
        \end{tabular}
    \label{tab:WMMA_vs_PTX}
\end{table}

@@

There are two main PTX instructions for using TCs. The first is \texttt{ldmatrix}, shown in Listing~\ref{lst:ldmatrix}, which creates a fragment of a matrix that is stored in registers shared across a warp. The second is \texttt{mma.sync}, shown in Listing~\ref{lst:mma}, which computes $\bm{D}=\bm{A}\times \bm{B}+\bm{C}$, where $\bm{A}$, $\bm{B}$, $\bm{C}$, and $\bm{D}$ are each fragments that were created by prior \texttt{ldmatrix} instructions. %The fragment $\bm{D}$ can be substituted for $\bm{C}$ to accumulate the result of the multiplication in place. 
When using FP16-32 instructions, the data stored in matrices $\bm{A}$ and $\bm{B}$ are represented in FP16 as 16$\times$16 and 16$\times$8 fragments, while the accumulator matrices, $\bm{C}$ and $\bm{D}$, are represented in FP32 as 16$\times$8 fragments.

@@

\noindent
{\bf Step~2:} Tensor cores operate on small matrices that are stored in registers; we refer to these matrices as ``register fragments,'' or ``rf''. There are three register fragments required to compute distances:
\begin{enumerate}[wide=0pt]
\item\prf: 16 dimensions from 16 points in dataset $D$. We refer to these simply as the ``points''. 
\item\qrf: 16 dimensions from 8 points in dataset $D$ in a transposed layout. We refer to these as the ``query points''. 
\item\arf: A $16\times 8$ fragment initialized with zeros.
 \end{enumerate}
 We use TCs to multiply and accumulate these three fragments to obtain: $\arf = \prf \times \qrf + \arf$. As TCs can only compute inner products of $16$ dimensions per operation (see Table~\ref{tab:WMMA_vs_PTX}), we must iteratively load new point data into \prf\ and \qrf\ and accumulate into \arf\ $\frac{d}{16}$ times.

@@

\begin{figure}
    \centering
    \includegraphics{tikz_build/Block_Tile}
    \caption{Two iterations of a $64{\times }64{\times }64$ block tile are shown. Two \sixfourd block fragments, \pbf \ and \qbf, are paged from global memory to shared memory (SMEM). Four warp tiles (Figure~\ref{fig:warp_mma}) each page a portion of the block fragments into their own register fragments and accumulate the distances. As $d{=}128$, this sequence is repeated for two iterations.}

@@

\subsubsection{L1 Cache elimination}\label{sec:L1_elimination}
Each SM has 192~KB of memory that can be used as an L1 cache or directly manipulated as shared memory. In order to maximize throughput, most of the L1 cache has been configured as shared memory to allow us to use a pipelined execution that hides memory access latency, which will be described in the following sections.

@@

Figure~\ref{fig:shared_memory} shows how each \eightd slice in a group of 8 threads will be stored in a unique bank and is thus free of bank conflicts.

While swizzling the data layout is not required to have coalesced and bank-conflict-free shared memory stores, it is imperative for loading the point data from shared memory and storing it into registers. To load a single fragment \prf, there are four phases, or memory transactions, in a \texttt{ldmatrix} instruction (see Listing~\ref{lst:ldmatrix} and Figure~\ref{fig:ldmatrix}). Each phase generates a single 128~B memory transaction that must be bank-conflict-free to achieve maximum throughput.

To see that banks are accessed in a conflict-free manner, observe that each \eightd slice of data being accessed in a single phase in Figure~\ref{fig:ldmatrix} is located in a unique bank in Figure~\ref{fig:shared_memory}. In contrast, a simple row-major copy from global memory (Figure~\ref{fig:global_memory}) into shared memory would create 8-way bank conflicts during each phase.

@@

We examine two main metrics when measuring performance. \emph{Response time:} We measure the response time of the algorithms. \emph{Derived TFLOPS:} We report the TFLOPS yielded by the TCs by taking the total number of operations and dividing it by the measured response time. We refer to this when we discuss TFLOPS.

@@

To make a fair comparison between all implementations, we include all overheads, such as allocating memory, transferring data to and from the GPU, and storing the results in main memory. Regarding methods that employ indexing, we include all index construction overheads.

@@

\noindent\textbf{\ouralg \ --} We configure \ouralg using the parameters outlined in Table~\ref{tab:optimized_parameters}, which were motivated in Section~\ref{sec:our_algorithm}.

\begin{table}[!t]
\footnotesize
\caption{Summary of optimized parameters. }
\begin{tabular}{p{5.4cm}|r} 
        \hline
        \rowcolor{blue!40!white}Parameter & Optimized Value \\ \hline
        Block tile dispatch shape & $8{\times }8$ blocks \\ \hline
        Block tile iteration size & $128{\times }128{\times }64$ \\ \hline
        Number of blocks in grid& $2{\times}$ \# of SMs (216 total) \\ \hline
        Warp tile iteration size & $64{\times }64{\times }16$ \\ \hline
        Warps per block & 4 \\ \hline
        Pipeline depth & 2 \\ \hline
        \end{tabular}
    \label{tab:optimized_parameters}
\end{table}

\noindent\textbf{\tedjoin \ --} As described in Section~\ref{sec:background_tedjoin}, \tedjoin~\cite{gallet2022leveraging} is the only other algorithm in the literature that computes Euclidean distances on TCs, and it uses double precision floating point (FP64) values.

@@

\noindent\textbf{\gdsjoin \ --} As described in Section~\ref{sec:related_work_gpu_similarity_searches} this algorithm performs index-supported range queries using CUDA cores. We use  the same optimizations and configuration as described in the \gdsjoin papers~\cite{gowanlock2019gpu,gowanlock2023optimization}, so we only outline the differences here. We configure \gdsjoin to use FP32 data, which provides a baseline of comparison to \ouralg that does not use FP64 like \tedjoin. Lastly, we use a batch size of $b_s=2\times10^9$ for batching the result set from the GPU to the host. We also employ \gdsjoin as executed using FP64 to compute the accuracy of \ouralg's mixed-precision results. The code is publicly available.\footnote{\url{https://github.com/mgowanlock/gpu_self_join}}

\noindent\textbf{\mistic \ --} This algorithm is similar to \gdsjoin above as it uses CUDA cores (Section~\ref{sec:related_work_gpu_similarity_searches}). We use the same configurations and optimizations as outlined in the \mistic paper~\cite{donnelly2024multi} and use a block size of 256 with 1024 blocks per kernel invocation (multiple invocations are used to batch the result set from device to host). The algorithm uses 6 levels and selects between 38 candidate layers at each level when incrementally constructing the index. \mistic is configured to use FP32 instead of FP64 (the latter was used in the evaluation in the \mistic paper~\cite{donnelly2024multi}). The code is publicly available.\footnote{\url{https://github.com/bwd29/self-join-MiSTIC}}

\noindent\textbf{Short circuiting of distance calculations -- }The CUDA core algorithms (\gdsjoin and \mistic) abort computing the distance between each pair of points when the running total distance exceeds $\epsilon$. When points in a dataset are very spread out, this saves significant computation time. The TC algorithms (\tedjoin and \ouralg) are less flexible because they compute the distances between all permutations of two sets of points simultaneously. \ouralg compares an entire 128$\times$128 block of points at a time. As \ouralg is computing 16,384 distances at a time, and every single point would need to short circuit for the optimization to be viable, we forego this optimization. \tedjoin compares an 8$\times$8 matrix of points at a time in FP64 mode, which is small enough that it does employ short circuiting. \tedjoinbrute does not abort early.

\begin{table}[!t]
\caption{Comparison of implementation properties.}
\footnotesize
    \centering
    {
        % Get Ted-Join-Brute to fit on one line
        % \setlength{\tabcolsep}{4pt} % default is 6pt
		\begin{tabular}{l|l|l|>{\raggedright\arraybackslash}p{1.2cm}|>{\raggedright\arraybackslash}p{1.8cm}} 
        \hline
        \rowcolor{blue!40!white}Implementation&GPU Core&Precision& Scenario~1: Brute Force  (Sec.~\ref{sec:scenario1_brute_with_epsilon})&Scenario~2: Index-Supported (Sec.~\ref{sec:scenario2_index})\\ \hline
        \ouralg&Tensor&FP16-32&\checkmark&\\
        \tedjoinbrute&Tensor&FP64&\checkmark&\\
        \tedjoinindex&Tensor&FP64&&\checkmark\\
        \gdsjoin&CUDA&FP32&&\checkmark\\
        \mistic & CUDA & FP32 & &\checkmark\\
        \hline
        \end{tabular}
    }
    \label{tab:implementations}
\end{table}

@@

\noindent\textbf{Selectivity of the Range Queries --} The search radius, $\epsilon$, used in a similarity search will impact the degree of pruning afforded by an indexing data structure. A large search radius that returns a large number of neighbors per point in a dataset will reduce the degree of index pruning, and a small value of $\epsilon$ will allow many points to be ignored when performing distance calculations. To standardize experiments across datasets, we employ the selectivity, which refers to the mean number of points found by each point searched in the dataset. The selectivity is defined as $S{=}(|R|{-}|D|)/|D|$, where $|R|$ is the total result set size (the number of pairs of points found within $\epsilon$ of each other). Common selectivity values are used to ensure that searches have a proportional amount of work to each other; otherwise, comparing performance between datasets is less meaningful. In our evaluation, we select values of $\epsilon$ for each dataset that yield three levels of selectivity (small, medium, and large), and these are defined as $S_s{=}64$, $S_m{=}128$, and $S_l{=}256$. The $\epsilon$ values corresponding to these selectivity values are given for the real-world datasets in Table~\ref{tab:datasets}.

\begin{table}[!t]
\caption{\emph{Top:} Real-world datasets that we use in our evaluation. The dataset size, $|D|$, and data dimensionality, $d$, are shown, in addition to the $\epsilon$ values that correspond to three target selectivity values that we employ ($S_{s}$,  $S_{m}$, and $S_{l}$). \emph{Bottom:} Synthetic datasets that we use in our evaluation.}
\footnotesize
    \centering
        \begin{tabular}{l|r|r|r|r|r} 
        \hline

            \rowcolor{blue!40!white} \multicolumn{6}{c}{Real-World Datasets}\\\hline
           \rowcolor{blue!40!white}  Dataset& $|D|$ & $d$  & $\epsilon (S_{s})$  &$\epsilon (S_{m})$  &$\epsilon (S_{l})$ \\\hline
            \sift&  10,000,000 &  128 & 122.5 & 136.5 & 152.5 \\ \hline
             \tinyds&  5,000,000 &  384 & 0.1831 & 0.2045 & 0.2275 \\ \hline
             \cifar&  60,000 &  512 & 0.6289 & 0.6591 & 0.6914 \\ \hline
             \gist&  1,000,000 &  960 & 0.4736 & 0.5292 & 0.5937 \\ \hline
           \rowcolor{blue!40!white}  \multicolumn{6}{c}{Synthetic Datasets}\\\hline
          \rowcolor{blue!40!white}   Dataset& \multicolumn{2}{c|}{ $|D|$ }& \multicolumn{3}{c}{$d$}\\\hline
             \synth&\multicolumn{2}{c|}{$10^{3 + n/3} \text{ where } n\in [0, 1, \ldots, 9]$}&\multicolumn{3}{c}{$d\in 2^n \text{ where } n\in[6, 7, \ldots, 12]$}\\\hline
        \end{tabular}
    \label{tab:datasets}
\end{table}

@@

Recall that because \ouralg is brute force, its performance is not impacted by the data distribution, so these results apply to any real-world dataset of the same size and dimensionality. To estimate \ouralg's performance on a dataset with a dimensionality that is not a multiple of $64$, one must use the throughput measurements from the next multiple of $64$. In other words, for low dimensionality like $d{=}65$, $63$ dimensions of zeros are added, and the throughput is almost halved. This effect on performance decreases as dimensionality increases.

@@

Table~\ref{tab:accuracy} presents the Overlap Between Result Sets metric across all real-world datasets, where we find that there is a negligible accuracy loss. The maximum accuracy loss occurs on \cifar ($S_l=256$), which still achieves a very high accuracy of 99.946\%.
